Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-18 latex/main.tex
Shorten section on daily sample cluster inference.

\subsubsection*{Tree inference from daily sample clusters}

With tens of thousands of samples being added to the ARG per day,
there are often clusters of hundreds of sequences attaching to the same node
(or more generally, the same recombinant path).
While some of these samples will require no extra mutations
(as they are identical to the attachment node),
in general there will be complex patterns of shared mutations among the samples
reflecting their evolutionary relationships.
A natural way to infer these within-cluster evolutionary relationships
is to use standard tree-building algorithms.
We can infer a plausible phylogeny relating the samples in a cluster
independently of the other samples in a daily batch
and then attach the phylogeny (and mutations) to the ARG at the node
identified by the LS model.

Sc2ts builds a local phylogeny relating the samples in each cluster as follows.
First, it infers a tree using a standard neighbor-joining
algorithm\cite{Saitou1987}. We first compute the Hamming distance between
all pairs of samples using scipy~\cite{Virtanen2020scipy},
and then use biotite\cite{Kunzmann2018biotite,Kunzmann2023biotite}
to compute the neighbor-joining tree.
Then, mutations are mapped back onto it using the Fitch-Hartigan parsimony 
algorithm \cite{Fitch1971toward,Hartigan1973minimum} implemented in tskit.
Finally, the resulting tree is added to the ARG at an attachment node
and the ARG is locally modified using heuristics to improve parsimony.
[TODO we need to define these?]

>>>>

\subsubsection*{Tree inference from daily sample clusters}
With tens of thousands of samples being added to the ARG per day,
there are often clusters of hundreds of sequences with the same
LS Viterbi solution. That is, large clusters of samples attach
to the same node LS copying (or more generally, the same recombinant path)
in the ARG. While some of these samples will require no extra mutations
(as they are identical to the attachment node),
in general there will be complex patterns of shared mutations among the samples
reflecting their evolutionary relationships.
We use a standard neighbor-joining algorithm\cite{Saitou1987}
to infer these within-cluster evolutionary relationships.
We first compute the Hamming distance between
all pairs of samples using scipy~\cite{Virtanen2020scipy},
and then use biotite\cite{Kunzmann2018biotite,Kunzmann2023biotite}
to compute the neighbor-joining tree.
Then, mutations are mapped back onto this daily sample cluster tree using 
the Fitch-Hartigan parsimony 
algorithm\cite{Fitch1971toward,Hartigan1973minimum} implemented in tskit.
Finally, the resulting tree and mutations are added to the ARG at the attachment node(s)
specified by the shared LS copying path.

\subsubsection*{Parsimony improving heuristics}
